Polynomial Time and Full Dependent Types

### 2023-07-11 paper/paper.tex
[ paper ] minor tweaks

\documentclass[acmsmall,review,screen,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=true,printacmref=true}

@@

As we saw above, contexts in QTT are comprised of variable
$\stackrel\rho:$ type triples, where $\rho$ is a natural number
indicating how many times the variable $x$ is available for use in a
$\sigma = 1$ term. There are two operations on raw contexts: scaling
$\pi\Gamma$, which multiplies each $\rho$ in $\Gamma$ by $\pi$, and
addition $\Gamma_1 + \Gamma_2$, which adds two contexts' usage
annotations assuming that the lengths and types are equal.  The basic
usage-annotation discipline of QTT is demonstrated by the context
formation and variable rules:
\begin{mathpar}
  \inferrule*
  { }
  {\epsilon~\isctxt}

@@

  \inferrule*
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}
\end{mathpar}
As with most dependent type theories, contexts are built inductively
from the empty context $\epsilon$ and extension of a context by a
variable with a type that is well-formed in the preceding
context. Usage annotations $\rho$ on variables are arbitrary, and
types are always judged in a $0$-annotated context. The variable rule
marks unused variables as usage $0$ and the selected variable with
usage $\sigma$.

\bob{Usage weakening somewhere}

% \begin{mathpar}

%   \inferrule* [right=Tm-Conv]
%   {\Gamma \vdash M \stackrel\sigma: S \\ 0\Gamma \vdash S \equiv T~\istype}
%   {\Gamma \vdash M \stackrel\sigma: T}
% \end{mathpar}

@@

\paragraph{Alternative definition} Every resource monoid induces a
partial ordering on its carrier set by $\alpha \leq \beta$ iff
$0 \leq M(\alpha, \beta)$. Taking this idea further, we can
reformulate a resource monoid as a symmetric monoidal category
enriched in the symmetric monoidal category $\natinf$, where the
monoid structure is addition. The conditions in the definition above
amount to the usual identity and composition laws for enriched
categories.

\paragraph{Agda Formalisation} Resource monoids are defined in the
module \AgdaModule{Algebra.ResourceMonoid}. We use a formulation
closer to the enriched category theory definition for the actual
formalisation, because it avoids having to treat equality in the
monoid structure separately from the induced preorder on elements.

@@

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. This construction is a
reformulation of \cite{hoffmann-dallago} suitable for modelling
dependent types. For a set $A$, the carrier of $L(A)$ is the set of
ternary relations $X \subseteq A \times M \times \ValSet$ and we
define the ordering $X \leq Y$ to hold iff there exists realising
expression $E \in \ExpSet$ and potential $\gamma \in M_0$ such that
for all $a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $e$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $e$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

@@

We have presented an two extensions of Quantitative Type Theory that
soundly and completely captures polynomial time.

@@

BLL, SAL, LAL

THe ones from Fossacs 2022, and POPL23(?) Romain Per...

@@

Blah

@@

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

>>>>

\documentclass[acmsmall,review,screen]{acmart}
\settopmatter{printfolios=true,printccs=true,printacmref=true}

@@

As we saw above, contexts in QTT are comprised of variable
$\stackrel\rho:$ type triples, where $\rho$ is a natural number
indicating how many times the variable $x$ is available for use in a
$\sigma = 1$ term. There are two operations on raw contexts: scaling
$\pi\Gamma$, which multiplies each $\rho$ in $\Gamma$ by $\pi$, and
addition $\Gamma_1 + \Gamma_2$, which adds two contexts' usage
annotations assuming that the lengths and types are equal. Contexts
are ordered pointwise $\Gamma' \sqsubseteq \Gamma$ on the usage
annotations (which is the \emph{reverse} ordering on naturals) The
basic usage-annotation discipline of QTT is demonstrated by the
context formation and variable rules:
\begin{mathpar}
  \inferrule*
  { }
  {\epsilon~\isctxt}

@@

  \inferrule*
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma'~\isctxt}
  {0\Gamma, x \stackrel\sigma: S, 0\Gamma' \vdash x \stackrel\sigma: S}

  \inferrule*
  {\Gamma \vdash M : S \\ \Gamma' \sqsubseteq \Gamma}
  {\Gamma' \vdash M : S}
\end{mathpar}
As with most dependent type theories, contexts are built inductively
from the empty context $\epsilon$ and extension of a context by a
variable with a type that is well-formed in the preceding
context. Usage annotations $\rho$ on variables are arbitrary, and
types are always judged in a $0$-annotated context. The variable rule
marks unused variables as usage $0$ and the selected variable with
usage $\sigma$.

@@

\paragraph{Alternative definition} Every resource monoid induces a
partial ordering on its carrier set by $\alpha \leq \beta$ iff
$0 \leq M(\alpha, \beta)$. Taking this idea further, we can
reformulate a resource monoid as a symmetric monoidal category
enriched in the symmetric monoidal category $\natinf$, where the
monoid structure is addition. The conditions in the definition above
amount to the usual identity and composition laws for enriched
categories. With this reading, we can see the value $M(\alpha,\beta)$
when it is $\geq 0$ as the possibility of moving from $\alpha$ to
$\beta$ levels of potential resource with some amount of residual
resource emitted for computation; when it is $-\infty$, moving from
$\alpha$ to $\beta$ is not possible.

\paragraph{Agda Formalisation} Resource monoids are defined in the
module \AgdaModule{Algebra.ResourceMonoid}. We use a formulation
closer to the enriched category theory definition for the actual
formalisation, because it avoids having to treat equality in the
monoid structure separately from the induced preorder on
elements. Thinking of proofs involving the resource monoid as a
process of finding a composable sequence of morphisms in a category
was a helpful intuition when constructing the realisability model
below.

@@

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. This construction is a
reformulation of \cite{hoffmann-dallago} suitable for modelling
dependent types. For a set $A$, the carrier of $L(A)$ is the set of
ternary relations $X \subseteq A \times M \times \ValSet$ and we
define the ordering $X \leq Y$ to hold iff there exists realising
expression $E \in \ExpSet$ and potential $\gamma \in M_0$ such that
for all $a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $E$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $E$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

@@

We have presented an two extensions of Quantitative Type Theory that
soundly and completely capture polynomial time. We have seen that this
allows for an expressive combination of verification and complexity
constrained computation. We now discuss some related work in implicit
computational complexity and linear dependent types, and take a look
at where the combination of resource conciousness and dependent types
could take us.

@@

BLL, SAL, LAL, LFPL

The ones from Fossacs 2022, and POPL23(?) Romain Per...

@@

@@

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}
